# Working with XRPD data in R {#intro}

At this stage it is assumed that you have R and RStudio installed on your computer and have installed the **powdR** package using `install.packages("powdR")`. For brief introductions to R, the [RStudio website has links to useful material](https://education.rstudio.com/learn/beginner/). One aspect worth noting is that any R function used throughout this document will have additional help associated with it that can be accessed using `?`. For example, to access help on the `mean()` function, use `?mean`.

This chapter will work through:

1. [The basic form of XRPD data](#basic-form)
2. [Loading XRPD data in R](#loading-data)
3. [Plotting XRPD data](#plotting-data)
4. [Manipulating XRPD data](#manipulating-data)

## The basic form of XRPD data {#basic-form}

In its most basic form, XRPD data (a diffractogram) is simply comprised of an x-axis (2θ and y-axis (counts). The following figure shows a diffratogram for a soil from north east Scotland.

```{r, fig.cap = "A diffractogram of a soil developed from Granite in north east Scotland", out.width='80%', fig.asp=.75, fig.align='center', echo = FALSE, message = FALSE, cache = TRUE}
library(powdR)

plot(soils$granite, wav = "Cu")
```

Each mineral contributing to the observed diffractogram contributes peaks at specific locations along the x-axis, each with characteristic relative intensities to one-another that are defined by the crystal structure and chemistry. Each mineral can therefore be considered to have a unique signature. For quartz, an omnipresent mineral in the world's soils, this signature is relatively simple:

```{r, fig.cap = "A diffractogram of quartz", out.width='80%', fig.asp=.75, fig.align='center', echo = FALSE, cache = TRUE}
plot(as_xy(data.frame(minerals$tth,
                      minerals$xrd$QUA.1)),
     wav = "Cu")
```

whereas other soil minerals such as K-feldspar have very different peak positions and relative intensities that again are governed by the mineral's crystal structure and chemistry:

```{r, fig.cap = "A diffractogram of K-feldspar (orthoclase)",out.width='80%', fig.asp=.75, fig.align='center', echo = FALSE, cache = TRUE}
plot(as_xy(data.frame(minerals$tth,
                      minerals$xrd$ORT)),
     wav = "Cu")
```

In any given diffractogram there can be multiple minerals contributing their unique signature to the observed pattern, which ends up simply being the sum of these signatures each scaled by different amounts (scaling factors). These scaling factors are inherently related to a given minerals concentration within the mixture based on the defined diffracting power of each mineral, known as Reference Intensity Ratios (see Chapter \@ref(quant)). Together these principles make XRPD the most widely used analytical technique for qualitative and quantitative soil mineralogy. 

## Loading XRPD data {#loading-data}

<p align="center"><iframe width="560" height="315" src="https://www.youtube.com/embed/0nO9fb0lcPQ" frameborder="0" allowfullscreen></iframe></p>

In order to work with XRPD data in R, it first needs to be loaded. XRPD data come in all sorts of proprietary formats (e.g. .raw, .dat and .xrdml), which can make this initial stage of loading data more complicated than it needs to be. As described above, XRPD data is most simply comprised of an x-axis (2θ) and y-axis (counts), and all XRPD data loaded into R throughout this documentation will hence take this XY form. Here 2 options for loading proprietary XRPD data into R will be described.

### Option 1: PowDLL {#powDLL}
The free software [PowDLL](http://users.uoi.gr/nkourkou/powdll/) written by Nikoloas Kourkoumelis, offers excellent functionality for the conversion of different XRPD file types. PowDLL can import and export a large range of XRPD file types including '.xy' files that can readily be loaded into R or any text editor. These '.xy' files are an ASCII format that simply comprises the two variables (2θ and counts) separated by a space. The following video from Phys Whiz on YouTube illustrates use of powDLL to create the '.xy' files that we seek to use.

<p align="center"><iframe width="560" height="315" src="https://www.youtube.com/embed/YUxg4-D-1Bk" frameborder="0" allowfullscreen></iframe></p>

Once you have your '.xy' files, they can be loaded into R using the `read_xy()` function from the **powdR** package. The following reproducible example uses files that are stored within **powdR** and were recorded on a Siemens D5000 using Co-K$\alpha$ radiation.

```{r, message=FALSE, warning=FALSE, cache = TRUE}
#load the powdR package
library(powdR)

#Extract the path of the file from the powdR package
file <- system.file("extdata/D5000/xy/D5000_1.xy", package = "powdR")

#Load the file as an object called xy1
xy1 <- read_xy(file)
```

```{r, message=FALSE, warning=FALSE, cache = TRUE}
#Explore the xy data
summary(xy1)

#check the class of xy data
class(xy1)
```

Notice how the class of `xy1` is both `XY` and `data.frame`. This means that various additional methods for each of these types of object classes can be used to explore and analyse the data. These methods can be viewed using:

```{r, message=FALSE, warning=FALSE, cache = TRUE}
methods(class = "XY")
```

which shows how functions [`align_xy()`](#alignment), [`interpolate()`](#interpolation) and [`plot()`](#plotting-data) all have methods for `XY` class objects. Help on each of these can be sourced using `?align_xy.XY`, `?interpolate.XY` and `?plot.XY`, respectively. When calling these functions it is not necessary to specify the `.XY` suffix because R will recognise the class and call the relevant method.

### Option 2: Loading directly into R {#loading-directly}
Alternatively to PowDLL, the `extract_xy()` function in the **powdR** package can extract the XY data from a wide range of proprietary XRPD file formats straight into R via the xylib C++ library implemented behind the scenes in the **rxylib** package [@rxylib2020].

```{r, message=FALSE, warning=FALSE, cache = TRUE}
#Extract the path of the file from the powdR package
file <- system.file("extdata/D5000/RAW/D5000_1.RAW", package = "powdR")

#Load the file as an object called xy2
xy2 <- extract_xy(file)

#Summarise the xy data
summary(xy2)

#Check the class of xy2
class(xy2)
```

A **word of warning** with `extract_xy()` is that it does not work with all proprietary file types. In particular you may experience problems with Bruker '.raw' files, in which case the use of PowDLL outlined [above](#powDLL) is recommended instead.

### Loading multiple files {#loading-multiple}
The two approaches for loading XRPD data outlined above can also be used to load any number of files into R at once. `read_xy()` and `extract_xy()` will recognise cases where more than one file path is supplied and therefore load the files into a `multiXY` object.

#### `read_xy()`
There are five '.xy' files stored within a directory of the **powdR** package that can be loaded into a `multiXY` object via:

```{r xy_list1, message=FALSE, warning=FALSE, cache = TRUE}
paths1 <- dir(system.file("extdata/D5000/xy", package = "powdR"),
             full.names = TRUE)

#Now read all files in the directory
xy_list1 <- read_xy(paths1)

#Check the class of xy_list1
class(xy_list1)
```

The resulting `multiXY` object is a list of `XY` objects, with each `XY` object being a data frame comprised of the 2θ and count intensities of the XRPD data.

```{r, message=FALSE, warning=FALSE, cache = TRUE}
#Check the class of each item within the multiXY object
lapply(xy_list1, class)
```

Each sample within the list can be accessed using the `$` symbol. For example:

```{r, message=FALSE, warning=FALSE, cache = TRUE}
#Summarise the data within the first sample:
summary(xy_list1$D5000_1)
```

Alternatively, the same item within `xy_list1` could be accessed using `xy_list1[[1]]`. In the same way the `XY` class objects have methods associated with them, there are a number of different methods for `multiXY` objects:

```{r, message=FALSE, warning=FALSE, cache = TRUE}
methods(class = "multiXY")
```

which include [`align_xy()`](#alignment), [`interpolate()`](interpolation), [`multi_xy_to_df()`](#to-data-frame) and [`plot`](#plotting) that are all detailed in subsequent sections.

#### `extract_xy()`
In addition to the five '.xy' files loaded above, there are also five '.RAW' files stored within a separate directory of **powdR**, which can be loaded in a similar fashion using `extract_xy()`:

```{r, message=FALSE, warning=FALSE, cache = TRUE}
paths2 <- dir(system.file("extdata/D5000/RAW", package = "powdR"),
              full.names = TRUE)

#Now read all files in the directory
xy_list2 <- extract_xy(paths2)

#Find out what the xy_list2 is
class(xy_list2)
```

which yields `xy_list2` that is identical to `xy_list1`:

```{r, message=FALSE, warning=FALSE, cache = TRUE}
all.equal(xy_list1, xy_list2)
```

## Plotting XRPD data {#plotting-data}

<p align="center"><iframe width="560" height="315" src="https://www.youtube.com/embed/XF3SaqeOAy0" frameborder="0" allowfullscreen></iframe></p>

The **powdR** package contains `plot()` methods for both `XY` and `multiXY` objects (see `?plot.XY` and `?plot.multiXY`).

### Plotting `XY` objects
An `XY` object can be plotted by:

```{r, fig.cap = "An example figure created using the plot method for an XY object.", message=FALSE, warning=FALSE, out.width='80%', fig.asp=.75, fig.align='center', cache = TRUE}
plot(xy1, wavelength = "Co", interactive = FALSE)
```

where `wavelength = "Co"` is required so that d-spacings can be computed and displayed when `interactive = TRUE`.

### Plotting `multiXY` objects
Often it's useful to plot more than one pattern at the same time, which can be achieved by plotting a `multiXY` object:

```{r, fig.cap = "An example figure created using the plot method for a multiXY object.", message=FALSE, warning=FALSE, out.width='80%', fig.asp=.75, fig.align='center', cache = TRUE}
plot(xy_list1, wavelength = "Co", interactive = FALSE)
```

As above, using `interactive = TRUE` in the function call will instead produce an interactive plot. In addition, the plotting of `XY` and `multiXY` objects also allows you to alter the x-axis limits and normalise the count intensities for easier comparison of specific peaks:

```{r, fig.cap = "An example figure created using the plot method for an XY object with normalised count intensities and a restrict x-axis.", message=FALSE, warning=FALSE, out.width='80%', fig.asp=.75, fig.align='center', cache = TRUE}
plot(xy_list1, wavelength = "Co",
     xlim = c(30, 32), normalise = TRUE)
```

### Modifying plots with **ggplot2** {#ggplot2-intro}
All plots shown so far are produced behind the scenes using the [**ggplot2**](https://ggplot2.tidyverse.org/) package, which will already be present on your machine if you have installed **powdR**. This means that it is possible to modify the plots in different ways by adding subsequent **ggplot2** layers, each separated by `+`. For example, it's possible to add points of the quartz peak intensities extracted from a [crystal structure database](http://rruff.geo.arizona.edu/AMS/amcsd.php) using `geom_point()`, and then add a title using `ggtitle()`, followed by changing the theme using `theme_bw()`.

```{r, fig.cap = "A quartz diffractogram with the locations and relative intensities of the quartz peaks identified.", message=FALSE, warning=FALSE, out.width='80%', fig.asp=.75, fig.align='center', cache = TRUE}
#Define the relative intensities of quartz peaks
quartz <- data.frame("tth" = c(24.22, 30.99, 42.61, 46.12,
                               47.06, 49.62, 53.62, 58.86,
                               64.60, 65.18, 70.79, 73.68),
                     "intensity" = c(0.20, 1.00, 0.06, 0.06,
                                     0.03, 0.05, 0.03, 0.11,
                                     0.03, 0.01, 0.07, 0.03))

#Load the ggplot2 package
library(ggplot2)

#Create a plot called p1
p1 <- plot(xy1, wav = "Co", normalise = TRUE) +
           geom_point(data = quartz, aes(x = tth, y = intensity), size = 5,
             shape = 21, colour = "red") +
           ggtitle("A soil with quartz peaks identified") +
           theme_bw()

p1
```

Further help on using the **ggplot2** package to build up plots in layers is provided in [Hadley Wickham's excellent documentation](https://r4ds.had.co.nz/data-visualisation.html) on data visualization.

Plots produced using **ggplot2** are static by default and can be exported as high quality images or pdfs. In some cases it is also useful to produce an interactive plot, which in the case of XRPD data allows for easy inspection of minor features. For most plots creating using **ggplot2**, the `ggplotly()` function from the **plotly** package can be used to convert them into interactive HTML plots that will load either in RStudio or your web browser:

```{r, eval = FALSE}
library(plotly)

ggplotly(p1)
```

## Manipulating XRPD data {#manipulating-data}

<p align="center"><iframe width="560" height="315" src="https://www.youtube.com/embed/pV8SdWDxMLQ" frameborder="0" allowfullscreen></iframe></p>

Loading XRPD data into R opens up almost limitless capabilities for analysing and manipulating the data via the R language and the [thousands of open source packages](https://cran.r-project.org/web/packages/available_packages_by_name.html) that enhance its functionality. Here some common forms of XRPD data manipulation will be introduced:

* [Subsetting](#subsetting)
* [Transformations of count intensities](#transformations)
* [Interpolation](#interpolation)
* [Alignment](#alignment)
* [Background fitting](#background)
* [Converting to data frames](#to-data-frame)
* [2θ transformation](#two-theta-transform)

### Subsetting XRPD data {#subsetting}
Quite often the analysis of XRPD data may be applied to a reduced 2θ range compared to that measured on the diffractometer. This can readily be achieved in R for any number of samples.

By summarising [`xy1`](#loading-directly) we can see that the 2θ column has a minimum of 2 and a maximum of 75 degrees.

```{r, message=FALSE, warning=FALSE, cache = TRUE}
summary(xy1)
```

If we wanted to reduce this to the range of 10--60 $^\circ$ 2θ then we have a couple of options. First we could extract the relevant data directly using the `[,]` notation for a data frame, where values before the comma represent rows and values after the comma represent columns:

```{r, message=FALSE, warning=FALSE, cache = TRUE}
xy1_sub <- xy1[xy1$tth >= 10 & xy1$tth <= 60, ]

summary(xy1_sub)
```

This first option is quite simple, but if we wanted to apply it to a list of patterns then using the `subset()` function would be a far better option.:

```{r, message=FALSE, warning=FALSE, cache = TRUE}
xy1_sub2 <- subset(xy1, tth >= 10 & tth <= 60)

identical(xy1_sub, xy1_sub2)
```

When using a function like `subset()`, it is very easy to apply it to any number of patterns in a `multiXY` object or `list` using `lapply()`:

```{r, message=FALSE, warning=FALSE, cache = TRUE}
xy_list1_sub <- lapply(xy_list1, subset,
                       tth >= 10 & tth <= 60)

#Similarly we can summarise the data in the list again
lapply(xy_list1_sub, summary)
```

### Transformations of count intensities {#transformations}
As will be introduced in subsequent chapters, log and root transforms of the count intensities of XRPD data can be useful when applying data mining or machine learning methods. By writing a function these transformations can be applied to any number of patterns in just a few lines of code:

```{r, fig.cap = "Log transformed XRPD data.", out.width='80%', fig.asp=.75, fig.align='center', message=FALSE, warning=FALSE, cache = TRUE}
#Create a function for log-transforming counts
log_xrpd <- function (x) {
  
  x[[2]] <- log10(x[[2]])
  
  return(x)
  
}

#apply the function to a list of XRPD data
xy_list1_log <- lapply(xy_list1,
                       log_xrpd)

#Plot the transformed data
plot(as_multi_xy(xy_list1_log), wavelength = "Cu")
```
Note how the use of `x[[2]]` in the function represents the second column of `x`. Alternatively, the form `x$counts` could be used but the function would fail to run if the variable name was altered in any way.

### Interpolation {#interpolation}
Sometimes XRPD patterns within a given data set may contain a number of different 2θ axes due to the measurements being carried out on different instruments or on the same instrument but with a different set-up. Direct comparison of such data requires that they are interpolated onto the same 2θ axis.

Here a data set containing 2 samples with different 2θ axes will be created using the `soils` and `rockjock_mixtures` data that are pre-loaded within the **powdR** package:

```{r, fig.cap = "Diffractograms from two different instruments.", out.width='80%', fig.asp=.75, fig.align='center', message=FALSE, warning=FALSE, cache = TRUE}
two_instruments <- as_multi_xy(list("a" = soils$granite,
                                    "b" = rockjock_mixtures$Mix2))

plot(two_instruments, wavelength = "Cu", normalise = TRUE)
```

In this example, the data within the `two_instruments` list will be interpolated onto an artificial 2θ axis called `new_tth`, which ranges from 10 to 60 $^\circ$ 2θ with a resolution of 0.02:

```{r, fig.cap = "Interpolated diffractograms from two different instruments.", out.width='80%', fig.asp=.75, fig.align='center', message=FALSE, warning=FALSE, cache = TRUE}
new_tth <- seq(10, 60, 0.02)

two_instruments_int <- interpolate(two_instruments, new_tth)

plot(two_instruments_int, wavelength = "Cu", normalise = TRUE)
```

### Alignment {#alignment}
Peak positions in XRPD data commonly shift in response to small variations in specimen height in the instrument. Even seemingly small misalignments between peaks can hinder the analysis of XRPD data [@Butler2019]. One approach to deal with such peak shifts is to use a mineral with essentially invariant peak positions as an internal standard (e.g. the common mineral quartz), resulting in well aligned data by adding or subtracting a fixed value to the 2θ axis.

The **powdR** package contains functionality for aligning single or multiple patterns using the `align_xy()` function. In the following examples, samples will be aligned to a pure quartz pattern that will be loaded from the **powdR** package using `read_xy()`

```{r, fig.cap = "Unaligned diffractograms.", out.width='80%', fig.asp=.75, fig.align='center', message=FALSE, warning=FALSE, cache = TRUE}
#Extract the location of the quartz xy file
quartz_file <- system.file("extdata/minerals/quartz.xy", package = "powdR")

#load the file
quartz <- read_xy(quartz_file)

#Plot the main quartz peak for pure quartz and a sandstone-derived soil
plot(as_multi_xy(list("quartz" = quartz,
                      "sandstone" = soils$sandstone)),
     wavelength = "Cu",
     normalise = TRUE,
     xlim = c(26, 27))
```

As shown in the figure above, the main quartz peaks of these two diffraction patterns do not align. This can be corrected using `align_xy()`:

```{r, fig.cap = "Aligned diffractograms.", out.width='80%', fig.asp=.75, fig.align='center', message=FALSE, warning=FALSE, cache = TRUE}
#Align the sandstone soil to the quartz pattern
sandstone_aligned <- align_xy(soils$sandstone, std = quartz,
                              xmin = 10, xmax = 60, xshift = 0.2)

#Plot the main quartz peak for pure quartz and a sandstone-derived soil
plot(as_multi_xy(list("quartz" = quartz,
                      "sandstone aligned" = sandstone_aligned)),
     wavelength = "Cu",
     normalise = TRUE,
     xlim = c(26, 27))
```

In cases where multiple patterns require alignment to a given standard, `align_xy()` can also be applied to `multiXY` objects:

```{r, fig.cap = "Unaligned diffractograms in a multiXY object.", out.width='80%', fig.asp=.75, fig.align='center', message=FALSE, warning=FALSE, cache = TRUE}
#Plot the unaligned soils data to show misalignments
plot(soils, wav = "Cu",
     xlim = c(26, 27), normalise = TRUE)
```

```{r, fig.cap = "Aligned diffractograms in a multiXY object.", out.width='80%', fig.asp=.75, fig.align='center', message=FALSE, warning=FALSE, cache = TRUE}
#Align the soils data to the quartz pattern
soils_aligned <- align_xy(soils, std = quartz,
                          xmin = 10, xmax = 60, xshift = 0.2)

#Plot the aligned data
plot(soils_aligned,
     wavelength = "Cu",
     normalise = TRUE,
     xlim = c(26, 27))
```

### Background fitting {#background}
Sometimes it is beneficial to fit and subtract the background from XRPD data. To achieve this, the **powdR** package includes the `bkg()` function which uses the peak filling method implemented in the **baseline** package [@baseline2010]. The fitting provided by the `bkg()` function uses four adjustable parameters that each have pre-loaded defaults (see `?bkg`):

```{r, fig.cap = "Fitting a background to a soil diffractogram. The y-axis is square root transformed to aid interpretation.", out.width='80%', fig.asp=.75, fig.align='center', message=FALSE, warning=FALSE, cache = TRUE}
#Fit a background to the sandstone-derived soil
granite_bkg <- bkg(soils$granite)

#summarise the resulting data
summary(granite_bkg)

#Plot the data and add a square root transform to aid interpretation
plot(granite_bkg) +
  scale_y_sqrt()
```

It is then simple to derive a background-subtracted `XY` object:

```{r, fig.cap = "Background subtracted diffractogram.", out.width='80%', fig.asp=.75, fig.align='center', message=FALSE, warning=FALSE, cache = TRUE}
#Calculate background subtracted count intensities
sandstone_bkg_sub <- as_xy(data.frame("tth" = granite_bkg$tth,
                                      "counts" = granite_bkg$counts - 
                                                 granite_bkg$background))

plot(sandstone_bkg_sub, wavelength = "Cu")
```

Sometimes the default values for the adjustable parameters (see `?bkg`) are not appropriate, and in such cases tuning of the parameters can be an iterative process. To help with this, there is a background fitting [Shiny](https://shiny.rstudio.com/) app that can be loaded from the **powdR** package using `run_bkg()`. This app will load in your default web browser and allows for the four adjustable parameters to be tuned to a given sample that is loaded into the app in '.xy' format.

### Converting to and from data frames {#to-data-frame}
`multiXY` objects can be converted to data frames using the `multi_xy_to_df()` function. When using this function, all samples within the `multiXY` object must be on the same 2θ axis, which can be ensured using the `interpolate()` function [outlined above](#interpolation).

```{r, message=FALSE, warning=FALSE, cache = TRUE}
#Convert xy_list1 to a dataframe
xy_df1 <- multi_xy_to_df(xy_list1, tth = TRUE)

#Show the first 6 rows of the derived data frame
head(xy_df1)
```
In cases where the 2θ column is not required, the use of `tth = FALSE` in the function call will result in only the count intensities being included in the output.

Data frames that take the form of `xy_df1` (i.e. that include the 2θ axis) can easily be converted back to a `multiXY` object using `as_multi_xy()`:

```{r, message=FALSE, warning=FALSE, cache = TRUE}
#Convert xy_df1 back to a multiXY list
back_to_list <- as_multi_xy(xy_df1)

#Check the class of the converted data
class(back_to_list)
```

### 2θ transformation {#two-theta-transform}
Laboratory XRPD data are usually collected using either Cu or Co X-ray tubes, which each have characteristic Kα wavelengths (e.g. Cu-Kα = 1.54056 Angstroms whereas Co-Kα = 1.78897 Angstroms). These wavelengths determine the 2θ at which the conditions for diffraction are met via Bragg's Law:

$$
\begin{aligned}
n\lambda = 2d\sin\theta
\end{aligned}
$$

where $n$ is an integer describing the diffraction order, $\lambda$ is the wavelength (Angstroms) and $d$ is the atomic spacing (Angstroms).

In some instances it can be useful to transform the 2θ axis of a given sample so that the 2θ peak positions are representative of a measurement made using a different X-ray source. This can be achieved using the `tth_transform()` function:

```{r, fig.cap = "Data obtained from Co and Cu X-ray tubes prior to 2theta transformation.", out.width='80%', fig.asp=.75, fig.align='center', message=FALSE, warning=FALSE, cache = TRUE}
#Create a multiXY object for this transform example
transform_eg <- as_multi_xy(list("Co" = xy_list1$D5000_1,
                                 "Cu" = soils$sandstone))

#Plot two patterns recorded using different wavelengths
plot(transform_eg,
     wavelength = "Cu",
     normalise = TRUE,
     interactive = FALSE)
```

```{r, fig.cap = "Data obtained from Co and Cu X-ray tubes after 2theta transformation.", out.width='80%', fig.asp=.75, fig.align='center', message=FALSE, warning=FALSE, cache = TRUE}
#transform the 2theta of the "Co" sample to "Cu"
transform_eg$Co$tth <- tth_transform(transform_eg$Co$tth,
                                     from = 1.78897,
                                     to = 1.54056)

#Replot data after transformation
plot(transform_eg,
     wavelength = "Cu",
     normalise = TRUE,
     interactive = FALSE)
```


Note how prior to the 2θ transformation, the dominant peaks in each pattern (associated with quartz in both cases) do not align. After the 2θ transformation the peaks are almost aligned, with a small additional 2θ shift that could be computed using the `align_xy()` function [outlined above](#alignment). Whilst Cu and Co are the most common X-ray sources for laboratory diffractometers, `tth_transform()` can accept any numeric wavelength value.
